\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-01-20 paper.qmd
Add model selection section

Visualization of NLDR techniques (m-vis) [@article59] for the same high-D data is particularly important in understanding and finding the best representation. If we have done so, the 2D models can be quite different from the data structure shown in high-D space. Therefore, visualizing the 2D model in high-D space (m-in-dis) is more useful to answer different types of questions:

@@

The process begins with a real data matrix $X$ containing $n$ observations in $p$ dimensions. It generates a sequence of $p$ × $d$ orthonormal projection matrices (bases), usually 1 or 2 dimensions. For each pair of orthonormal bases $A_t$ and $A_{t+1}$, a geodesic path is interpolated to create smooth animation between projections.

@@

@@

num_bins_for_nldr_methods_df <- tibble::tibble(NLDR_technique = "tSNE", tot_bins = 8)

num_bins_for_nldr_methods_df <- num_bins_for_nldr_methods_df |> 
  tibble::add_row(NLDR_technique = "UMAP", tot_bins = 6)

num_bins_for_nldr_methods_df <- num_bins_for_nldr_methods_df |> 
  tibble::add_row(NLDR_technique = "PHATE", tot_bins = 8)

num_bins_for_nldr_methods_df <- num_bins_for_nldr_methods_df |> 
  tibble::add_row(NLDR_technique = "TriMAP", tot_bins = 6)

num_bins_for_nldr_methods_df <- num_bins_for_nldr_methods_df |> 
  tibble::add_row(NLDR_technique = "PaCMAP", tot_bins = 10)

# knitr::kable(
#   num_bins_for_nldr_methods_df,
#   col.names = c('NLDR technique', 'total number of bins'), format = "latex", 
#              caption = "Total number of bins suggested by rule of thumb for each NLDR technique applied to S-curve dataset."
# ) |> 
#   kable_styling(latex_options = c("striped", "HOLD_position"), position = "center", full_width = FALSE) |>
#   row_spec(0, bold = TRUE)
num_bins_for_nldr_methods_df$tot_bins <- as.character(num_bins_for_nldr_methods_df$tot_bins)
colnames(num_bins_for_nldr_methods_df) <- c('NLDR technique', 'total number of bins')
num_bins_for_nldr_methods_df <- tibble::remove_rownames(num_bins_for_nldr_methods_df)

xtable(num_bins_for_nldr_methods_df, caption = "Total number of bins suggested by rule of thumb for each NLDR technique applied to S-curve dataset.", label = "tbl-msebinsnldr") |>
  xtable2kable() |> 
  kable_styling(latex_options = c("striped", "HOLD_position"), position = "center", full_width = FALSE) |>
  row_spec(0, bold = TRUE)

@@

@@

@@

@@

```{r}
#| echo: false
#| warning: false
#| fig-cap: Goodness of fit statistics from different NLDR techniques applied to training S-curve dataset with different benchmark values to remove the low-density heaxgons. What is the effective benchmark value to remove the low-density heaxgons?
#| label: fig-diagnosticpltScurvelwd
##| out-width: 100%

@@

There is no definitive rule for determining the benchmark value to remove long edges. However, we have a method to find a default value (see Appendix). Adjusting values around this default can help to find benchmark value to remove long edges, contributing to the construction of a smoother 2D representation.

@@

### Model selection {#sec-selection}

- By comaring AIC what is the best NLDR technique to reprrsnt NLDR data

>>>>

Visualization of NLDR techniques (m-vis) [@article59] for the same high-D data is particularly important in understanding and finding the best representation. If we have done so, the 2D models can be quite different from the data structure shown in high-D space. Therefore, visualizing the 2D model in high-D space (m-in-ds) is more useful to answer different types of questions:

@@

The process begins with a real data matrix $X$ containing $n$ observations in $p$ dimensions. It generates a sequence of $p$ × $d$ orthonormal projection matrices (bases) $A_t$, usually $d$ is $1$ or $2$ dimensions. For each pair of orthonormal bases $A_t$ and $A_{t+1}$, a geodesic path is interpolated to create smooth animation between projections.

@@

In this paper, we introduce a novel approach, which is designed to determine which method, and which parameter choice provide the most useful representation of high dimensional data. Our approach involves dividing the data set into two subsets: a training set, which serves to construct the low-dimensional manifold, and a test set, utilized to generate predictive values and residuals. For the implementation of our approach, a 2D embedding data set serves as the initiation point for the subsequent analysis and visualization. The output

write about the S-curve as well.

@@

num_bins_for_nldr_methods_df <- tibble::tibble(NLDR_technique = "tSNE", tot_bins = 72, aic = -21882.462)

num_bins_for_nldr_methods_df <- num_bins_for_nldr_methods_df |> 
  tibble::add_row(NLDR_technique = "UMAP", tot_bins = 98, aic = -20013.266)

num_bins_for_nldr_methods_df <- num_bins_for_nldr_methods_df |> 
  tibble::add_row(NLDR_technique = "PHATE", tot_bins = 90, aic = -15036.91)

num_bins_for_nldr_methods_df <- num_bins_for_nldr_methods_df |> 
  tibble::add_row(NLDR_technique = "TriMAP", tot_bins = 154, aic = -23001.90)

num_bins_for_nldr_methods_df <- num_bins_for_nldr_methods_df |> 
  tibble::add_row(NLDR_technique = "PaCMAP", tot_bins = 132, aic = -21330.059)

# knitr::kable(
#   num_bins_for_nldr_methods_df,
#   col.names = c('NLDR technique', 'total number of bins'), format = "latex", 
#              caption = "Total number of bins suggested by rule of thumb for each NLDR technique applied to S-curve dataset."
# ) |> 
#   kable_styling(latex_options = c("striped", "HOLD_position"), position = "center", full_width = FALSE) |>
#   row_spec(0, bold = TRUE)
num_bins_for_nldr_methods_df$tot_bins <- as.character(num_bins_for_nldr_methods_df$tot_bins)
colnames(num_bins_for_nldr_methods_df) <- c('NLDR technique', 'total number of bins', "AIC")
num_bins_for_nldr_methods_df <- tibble::remove_rownames(num_bins_for_nldr_methods_df)

xtable(num_bins_for_nldr_methods_df, caption = "Total number of bins suggested by rule of thumb and AIC values for each NLDR technique applied to S-curve dataset.", label = "tbl-msebinsnldr") |>
  xtable2kable() |> 
  kable_styling(latex_options = c("striped", "HOLD_position"), position = "center", full_width = FALSE) |>
  row_spec(0, bold = TRUE) |>
  row_spec(4, bold=T, color = "red")

@@

vec <- stats::setNames(rep("", 5), c("number_of_bins", "number_of_observations", "total_error", "total_mse", "num_bins_x"))  ## Define column names

@@

  eval_df_training <- eval_df_training |>
    mutate(num_bins_x = num_bins_vec[i])

@@

```{r}
#| echo: false
#| message: false
#| warning: false

hex_full_count_df_loop2 <- hex_full_count_df_loop2 |>
  dplyr::mutate(type = if_else(std_counts < 0.12, "low", "high"))

hex_full_count_df_filtered <- hex_full_count_df_loop2 |>
  dplyr::filter(type == "low")

p2_n <-  ggplot(data = hex_full_count_df_loop2, aes(x = x, y = y)) +
  geom_polygon(color = "black", aes(group = polygon_id, fill = std_counts)) +
  #geom_point(data = UMAP_data, aes(x = UMAP1, y = UMAP2), alpha = 0.5) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  #coord_cartesian(xlim =c(-5, 7), ylim = c(-10, 10)) +
  theme_void() +
  theme(legend.position="none", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6)) +
  guides(fill = guide_colourbar(title = "Standardized count")) #+
  #annotate(geom = 'text', label = "b", x = -Inf, y = Inf, hjust = -0.3, vjust = 1, size = 3)

p2_n_rm <-  ggplot(data = hex_full_count_df_loop2, aes(x = x, y = y)) +
  geom_polygon(color = "black", aes(group = polygon_id, fill = std_counts)) +
  geom_text(data = hex_full_count_df_filtered, color = "red", aes(x = c_x, y = c_y, group = polygon_id, label = "X", size = 0.5)) +
  #geom_point(data = UMAP_data, aes(x = UMAP1, y = UMAP2), alpha = 0.5) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  #coord_cartesian(xlim =c(-5, 7), ylim = c(-10, 10)) +
  theme_void() +
  theme(legend.position="none", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6)) +
  guides(fill = guide_colourbar(title = "Standardized count")) #+
  #annotate(geom = 'text', label = "b", x = -Inf, y = Inf, hjust = -0.3, vjust = 1, size = 3)

```{r}
#| echo: false
#| warning: false
#| fig-cap: Goodness of f
#| label: fig-rmlowdenshex
##| out-width: 100%

p2_n + p2_n_rm +
  plot_annotation(tag_levels = 'a') +
  plot_layout(guides='collect', ncol = 2) &
  theme(legend.position='bottom', plot.tag = element_text(size = 8))
```

@@

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Goodness of fit statistics from different NLDR techniques applied to training S-curve dataset with different benchmark values to remove the low-density heaxgons. What is the effective benchmark value to remove the low-density heaxgons?"
#| label: fig-diagnosticpltScurvelwd
##| out-width: 100%

@@

There is no definitive rule for determining the benchmark value to remove long edges. However, we have a method to find a default value (see Appendix). Adjusting values around this default can help to find benchmark value to remove long edges, contributing to the construction of a smoother 2D representation (see @fig-diagnosticpltScurvelgrm).

@@

#### Model selection {#sec-selection}

In the process of selecting the most suitable NLDR technique to represent high-D data, the Akaike Information Criterion (AIC) stands as a crucial tool for insightful comparisons. Once the effective number of bins has been determined for each method, AIC values can be compared to identify the best NLDR technique with the lowest AIC value. For example, as shown in @fig-diagnosticpltScurve (a), TriMAP is the most suitable NLDR technique to represent the S-curve data (see table 1).

### 2024-01-20 paper.tex
Add model selection section

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, breakable]}{\end{tcolorbox}}\fi

@@

Visualization of NLDR techniques (m-vis) \citep{article59} for the same
high-D data is particularly important in understanding and finding the
best representation. If we have done so, the 2D models can be quite
different from the data structure shown in high-D space. Therefore,
visualizing the 2D model in high-D space (m-in-dis) is more useful to
answer different types of questions:

@@

The process begins with a real data matrix \(X\) containing \(n\)
observations in \(p\) dimensions. It generates a sequence of \(p\) ×
\(d\) orthonormal projection matrices (bases), usually 1 or 2
dimensions. For each pair of orthonormal bases \(A_t\) and \(A_{t+1}\),
a geodesic path is interpolated to create smooth animation between
projections.

@@

\begin{table}[H]
\centering
\begin{tabular}{rll}
  \hline
\textbf{} & \textbf{NLDR technique} & \textbf{total number of bins }\\ 
  \hline
\cellcolor{gray!6}{1} & \cellcolor{gray!6}{tSNE} & \cellcolor{gray!6}{8 }\\ 
  2 & UMAP & 6 \\ 
\cellcolor{gray!6}{  3} & \cellcolor{gray!6}{PHATE} & \cellcolor{gray!6}{8 }\\ 
  4 & TriMAP & 6 \\ 
\cellcolor{gray!6}{  5} & \cellcolor{gray!6}{PaCMAP} & \cellcolor{gray!6}{10 }\\ 
   \hline
\end{tabular}
\caption{Total number of bins suggested by rule of thumb for each NLDR technique applied to S-curve dataset.} 
\label{tbl-msebinsnldr}
\end{table}

@@

@@

There is no definitive rule for determining the benchmark value to
remove long edges. However, we have a method to find a default value
(see Appendix). Adjusting values around this default can help to find
benchmark value to remove long edges, contributing to the construction
of a smoother 2D representation.

@@

\hypertarget{sec-selection}{%
\subsection{Model selection}\label{sec-selection}}

\begin{itemize}
\tightlist
\item
  By comaring AIC what is the best NLDR technique to reprrsnt NLDR data
\end{itemize}

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, frame hidden, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, sharp corners]}{\end{tcolorbox}}\fi

@@

Visualization of NLDR techniques (m-vis) \citep{article59} for the same
high-D data is particularly important in understanding and finding the
best representation. If we have done so, the 2D models can be quite
different from the data structure shown in high-D space. Therefore,
visualizing the 2D model in high-D space (m-in-ds) is more useful to
answer different types of questions:

@@

The process begins with a real data matrix \(X\) containing \(n\)
observations in \(p\) dimensions. It generates a sequence of \(p\) ×
\(d\) orthonormal projection matrices (bases) \(A_t\), usually \(d\) is
\(1\) or \(2\) dimensions. For each pair of orthonormal bases \(A_t\)
and \(A_{t+1}\), a geodesic path is interpolated to create smooth
animation between projections.

@@

\begin{table}[H]
\centering
\begin{tabular}{rllr}
  \hline
\textbf{} & \textbf{NLDR technique} & \textbf{total number of bins} & \textbf{AIC }\\ 
  \hline
\cellcolor{gray!6}{1} & \cellcolor{gray!6}{tSNE} & \cellcolor{gray!6}{72} & \cellcolor{gray!6}{-21882.46 }\\ 
  2 & UMAP & 98 & -20013.27 \\ 
\cellcolor{gray!6}{  3} & \cellcolor{gray!6}{PHATE} & \cellcolor{gray!6}{90} & \cellcolor{gray!6}{-15036.91 }\\ 
\textcolor{red}{\textbf{  4}} & \textcolor{red}{\textbf{TriMAP}} & \textcolor{red}{\textbf{154}} & \textcolor{red}{\textbf{-23001.90 }}\\ 
\cellcolor{gray!6}{  5} & \cellcolor{gray!6}{PaCMAP} & \cellcolor{gray!6}{132} & \cellcolor{gray!6}{-21330.06 }\\ 
   \hline
\end{tabular}
\caption{Total number of bins suggested by rule of thumb and AIC values for each NLDR technique applied to S-curve dataset.} 
\label{tbl-msebinsnldr}
\end{table}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-rmlowdenshex-1.pdf}

\caption{\label{fig-rmlowdenshex}Goodness of f}

@@

There is no definitive rule for determining the benchmark value to
remove long edges. However, we have a method to find a default value
(see Appendix). Adjusting values around this default can help to find
benchmark value to remove long edges, contributing to the construction
of a smoother 2D representation (see
Figure~\ref{fig-diagnosticpltScurvelgrm}).

@@

\hypertarget{sec-selection}{%
\subsubsection{Model selection}\label{sec-selection}}

In the process of selecting the most suitable NLDR technique to
represent high-D data, the Akaike Information Criterion (AIC) stands as
a crucial tool for insightful comparisons. Once the effective number of
bins has been determined for each method, AIC values can be compared to
identify the best NLDR technique with the lowest AIC value. For example,
as shown in Figure~\ref{fig-diagnosticpltScurve} (a), TriMAP is the most
suitable NLDR technique to represent the S-curve data (see table 1).
